### 2024-03-03 presentation/parts/overview_presentation.tex
WIP

\newcommand{\tikzoverview}{%
    \resizebox{!}{0.8\textheight}{%
        \begin{tikzpicture}[x=1cm,y=1cm]
            \node [mybox, fill=presentation_orange] at (0,5) (predictions) {};
            \foreach \i in {-11,...,11} {
                \fill[myrect, fill=presentation_orange_shaded] (0.2*\i,4.8) rectangle ++(0.1,0.4);
            }
            \node [] at (0,5) {\small\textsc{predictions}};

@@

\newcommand{\tikzhierarchical}{%
    \resizebox{!}{0.8\textheight}{%
        \begin{tikzpicture}[x=1cm,y=1cm]
            \node [nearly transparent, mybox, fill=presentation_orange] at (0,5) (predictions) {};
            \foreach \i in {-11,...,11} {
                \fill[nearly transparent, myrect, fill=presentation_orange_shaded] (0.2*\i,4.8) rectangle ++(0.1,0.4);
            }
            \node [nearly transparent] at (0,5) {\small\textsc{predictions}};

@@

\newcommand{\tikzmodelagnostic}{%
    \resizebox{!}{0.8\textheight}{%
        \begin{tikzpicture}[x=1cm,y=1cm]
            \node [nearly transparent, mybox, fill=presentation_orange] at (0,5) (predictions) {};
            \foreach \i in {-11,...,11} {
                \fill[nearly transparent, myrect, fill=presentation_orange_shaded] (0.2*\i,4.8) rectangle ++(0.1,0.4);
            }
            \node [nearly transparent] at (0,5) {\small\textsc{predictions}};

@@

\newcommand{\tikzbrief}{%
    \resizebox{!}{0.8\textheight}{%
        \begin{tikzpicture}[x=1cm,y=1cm]
            \node [nearly transparent, mybox, fill=presentation_orange] at (0,5) (predictions) {};
            \foreach \i in {-11,...,11} {
                \fill[nearly transparent, myrect, fill=presentation_orange_shaded] (0.2*\i,4.8) rectangle ++(0.1,0.4);
            }
            \node [nearly transparent] at (0,5) {\small\textsc{predictions}};

@@

\newcommand{\tikzbenchmarking}{%
    \resizebox{!}{0.8\textheight}{%
        \begin{tikzpicture}[x=1cm,y=1cm]
            \node [nearly transparent, mybox, fill=presentation_orange] at (0,5) (predictions) {};
            \foreach \i in {-11,...,11} {
                \fill[nearly transparent, myrect, fill=presentation_orange_shaded] (0.2*\i,4.8) rectangle ++(0.1,0.4);
            }
            \node [nearly transparent] at (0,5) {\small\textsc{predictions}};

@@

\newcommand{\tikzautomated}{%
    \resizebox{!}{0.8\textheight}{%
        \begin{tikzpicture}[x=1cm,y=1cm]
            \node [mybox, fill=presentation_orange] at (0,5) (predictions) {};
            \foreach \i in {-11,...,11} {
                \fill[myrect, fill=presentation_orange_shaded] (0.2*\i,4.8) rectangle ++(0.1,0.4);
            }
            \node [] at (0,5) {\small\textsc{predictions}};

@@

\newcommand{\tikzretrospective}{%
    \resizebox{!}{0.8\textheight}{%
        \begin{tikzpicture}[x=1cm,y=1cm]
            \node [mybox, fill=presentation_orange] at (0,5) (predictions) {};
            \foreach \i in {-11,...,11} {
                \fill[myrect, fill=presentation_orange_shaded] (0.2*\i,4.8) rectangle ++(0.1,0.4);
            }
            \node [] at (0,5) {\small\textsc{predictions}};

@@

                    \tikzmarkin<3>{modelagnostic-1}(0.1,-0.4)(-0.1,0.6)
                    & {\color<2->{black}\makecell[l]{\textsc{\itshape chapter 5}\\\\}}     & {\color<2->{black} \bfseries \scshape \Large \makecell[l]{model-agnostic out-of-distribution detection\\ using combined statistical tests}} & \tikzmarkend{modelagnostic-1} \\
                    \addlinespace[0.5em]
                    \addlinespace[0.5em]

                    \tikzmarkin<4>{brief-1}(0.1,-0.4)(-0.1,0.6)
                    & {\color<2->{black}\makecell[l]{ \textsc{\itshape chapter 6}\\\\}}    & {\color<2->{black} \bfseries \scshape \Large \makecell[l]{a brief overview of unsupervised speech \\representation learning}} & \tikzmarkend{brief-1} \\
                    \addlinespace[0.5em]
                    \addlinespace[0.5em]

                    \tikzmarkin<5>{benchmarking-1}(0.1,-0.3)(-0.1,0.4)
                    & {\color<2->{black}\makecell[l]{\textsc{\itshape chapter 7}}}         & {\color<2->{black} \bfseries \scshape \Large \makecell[l]{benchmarking latent variable models for speech}} & \tikzmarkend{benchmarking-1} \\
                    \addlinespace[0.5em]
                    \addlinespace[0.5em]

                    \tikzmarkin<6>{automated-1}(0.1,-0.4)(-0.1,0.6)
                    & {\color<2->{black}\makecell[l]{\textsc{\itshape chapter 8}\\\\}}     & {\color<2->{black} \bfseries \scshape \Large \makecell[l]{automated medical coding on mimic-iii and \\mimic-iv: a critical review and replicability study}} & \tikzmarkend{automated-1} \\
                    \addlinespace[0.5em]
                    \addlinespace[0.5em]

                    \tikzmarkin<7>{retrospective-1}(0.1,-0.4)(-0.1,0.6)
                    & {\color<2->{black}\makecell[l]{\textsc{\itshape chapter 9}\\\\}}     & {\color<2->{black} \bfseries \scshape \Large \makecell[l]{a retrospective study on machine learning-\\assisted stroke recognition for medical helpline calls}} & \tikzmarkend{retrospective-1} \\
                    \addlinespace[0.5em]
                    \midrule
                    \addlinespace[0.5em]

                    & {\color<2->{black!20}\makecell[l]{\textsc{\itshape chapter 10}}}        & {\color<2->{black!20} \bfseries \scshape \Large \makecell[l]{discussion and conclusion}} & \\
                \end{tabular}
                }
            \end{table}
        \end{column}
        \begin{column}{0.4\textwidth}
            \tikzstyle{mybox} = [fill, rounded corners, minimum width=140, minimum height=20]
            \tikzstyle{myrect} = [fill]
            \begin{figure}
                \centering
                \begin{overprint}
                    \onslide<1>\tikzoverview
                    \onslide<2>\tikzhierarchical
                    \onslide<3>\tikzmodelagnostic
                    \onslide<4>\tikzbrief
                    \onslide<5>\tikzbenchmarking
                    \onslide<6>\tikzautomated
                    \onslide<7>\tikzretrospective
                \end{overprint}
            \end{figure}
        \end{column}
    \end{columns}
\end{frame}

>>>>

\newcommand{\tikzoverview}{%
    \resizebox{0.8\textwidth}{!}{%
        \begin{tikzpicture}[x=1cm,y=1cm]
            \node [mybox, fill=presentation_orange] at (0,5) (predictions) {};
            \foreach \i in {-11,...,11} {
                \fill[myrect, fill=presentation_orange_shaded] (0.2*\i,4.8) rectangle ++(0.1,0.4);
            }
            \node [] at (0,5) {\small\textsc{predictions}};

@@

\newcommand{\tikzhierarchical}{%
    \resizebox{0.8\textwidth}{!}{%
        \begin{tikzpicture}[x=1cm,y=1cm]
            \node [nearly transparent, mybox, fill=presentation_orange] at (0,5) (predictions) {};
            \foreach \i in {-11,...,11} {
                \fill[nearly transparent, myrect, fill=presentation_orange_shaded] (0.2*\i,4.8) rectangle ++(0.1,0.4);
            }
            \node [nearly transparent] at (0,5) {\small\textsc{predictions}};

@@

\newcommand{\tikzmodelagnostic}{%
    \resizebox{0.8\textwidth}{!}{%
        \begin{tikzpicture}[x=1cm,y=1cm]
            \node [nearly transparent, mybox, fill=presentation_orange] at (0,5) (predictions) {};
            \foreach \i in {-11,...,11} {
                \fill[nearly transparent, myrect, fill=presentation_orange_shaded] (0.2*\i,4.8) rectangle ++(0.1,0.4);
            }
            \node [nearly transparent] at (0,5) {\small\textsc{predictions}};

@@

\newcommand{\tikzbrief}{%
    \resizebox{0.8\textwidth}{!}{%
        \begin{tikzpicture}[x=1cm,y=1cm]
            \node [nearly transparent, mybox, fill=presentation_orange] at (0,5) (predictions) {};
            \foreach \i in {-11,...,11} {
                \fill[nearly transparent, myrect, fill=presentation_orange_shaded] (0.2*\i,4.8) rectangle ++(0.1,0.4);
            }
            \node [nearly transparent] at (0,5) {\small\textsc{predictions}};

@@

\newcommand{\tikzbenchmarking}{%
    \resizebox{0.8\textwidth}{!}{%
        \begin{tikzpicture}[x=1cm,y=1cm]
            \node [nearly transparent, mybox, fill=presentation_orange] at (0,5) (predictions) {};
            \foreach \i in {-11,...,11} {
                \fill[nearly transparent, myrect, fill=presentation_orange_shaded] (0.2*\i,4.8) rectangle ++(0.1,0.4);
            }
            \node [nearly transparent] at (0,5) {\small\textsc{predictions}};

@@

\newcommand{\tikzautomated}{%
    \resizebox{0.8\textwidth}{!}{%
        \begin{tikzpicture}[x=1cm,y=1cm]
            \node [mybox, fill=presentation_orange] at (0,5) (predictions) {};
            \foreach \i in {-11,...,11} {
                \fill[myrect, fill=presentation_orange_shaded] (0.2*\i,4.8) rectangle ++(0.1,0.4);
            }
            \node [] at (0,5) {\small\textsc{predictions}};

@@

\newcommand{\tikzretrospective}{%
    \resizebox{0.8\textwidth}{!}{%
        \begin{tikzpicture}[x=1cm,y=1cm]
            \node [mybox, fill=presentation_orange] at (0,5) (predictions) {};
            \foreach \i in {-11,...,11} {
                \fill[myrect, fill=presentation_orange_shaded] (0.2*\i,4.8) rectangle ++(0.1,0.4);
            }
            \node [] at (0,5) {\small\textsc{predictions}};

@@

                    \tikzmarkin<2>{modelagnostic-1}(0.1,-0.4)(-0.1,0.6)
                    & {\color<2->{black}\makecell[l]{\textsc{\itshape chapter 5}\\\\}}     & {\color<2->{black} \bfseries \scshape \Large \makecell[l]{model-agnostic out-of-distribution detection\\ using combined statistical tests}} & \tikzmarkend{modelagnostic-1} \\
                    \addlinespace[0.5em]
                    \addlinespace[0.5em]

                    \tikzmarkin<3>{brief-1}(0.1,-0.4)(-0.1,0.6)
                    & {\color<2->{black}\makecell[l]{ \textsc{\itshape chapter 6}\\\\}}    & {\color<2->{black} \bfseries \scshape \Large \makecell[l]{a brief overview of unsupervised speech \\representation learning}} & \tikzmarkend{brief-1} \\
                    \addlinespace[0.5em]
                    \addlinespace[0.5em]

                    \tikzmarkin<4>{benchmarking-1}(0.1,-0.3)(-0.1,0.4)
                    & {\color<2->{black}\makecell[l]{\textsc{\itshape chapter 7}}}         & {\color<2->{black} \bfseries \scshape \Large \makecell[l]{benchmarking latent variable models for speech}} & \tikzmarkend{benchmarking-1} \\
                    \addlinespace[0.5em]
                    \addlinespace[0.5em]

                    \tikzmarkin<5>{automated-1}(0.1,-0.4)(-0.1,0.6)
                    & {\color<2->{black}\makecell[l]{\textsc{\itshape chapter 8}\\\\}}     & {\color<2->{black} \bfseries \scshape \Large \makecell[l]{automated medical coding on mimic-iii and \\mimic-iv: a critical review and replicability study}} & \tikzmarkend{automated-1} \\
                    \addlinespace[0.5em]
                    \addlinespace[0.5em]

                    \tikzmarkin<6>{retrospective-1}(0.1,-0.4)(-0.1,0.6)
                    & {\color<2->{black}\makecell[l]{\textsc{\itshape chapter 9}\\\\}}     & {\color<2->{black} \bfseries \scshape \Large \makecell[l]{a retrospective study on machine learning-\\assisted stroke recognition for medical helpline calls}} & \tikzmarkend{retrospective-1} \\
                    \addlinespace[0.5em]
                    \midrule
                    \addlinespace[0.5em]

                    & {\color<2->{black!20}\makecell[l]{\textsc{\itshape chapter 10}}}        & {\color<2->{black!20} \bfseries \scshape \Large \makecell[l]{discussion and conclusion}} & \\
                \end{tabular}
                }
            \end{table}
        \end{column}
        \begin{column}{0.4\textwidth}
            \begin{figure}
                \centering
                \begin{overprint}
                    \onslide<1>\tikzoverview
                    \onslide<2>\tikzhierarchical
                    % \onslide<2>\tikzmodelagnostic
                    \onslide<3>\tikzbrief
                    \onslide<4>\tikzbenchmarking
                    \onslide<5>\tikzautomated
                    \onslide<6>\tikzretrospective
                \end{overprint}
            \end{figure}
        \end{column}
    \end{columns}

    \note[item]<1->{So how does the thesis tackle this problem of building such a system?}
    \note[item]<2->{First we examine how a certain class of generative models can be used to detect data that is different from training data.}
    \note[item]<2->{In our second paper we generalize this to be model-agnostic and phrased as a statistical test.}
    \note[item]<3->{We then take a step back and provide an overview of modern methods for unsupervised speech representation learning.}
    \note[item]<4->{We then benchmark some of these methods, generative latent variable models, on likelihood and automatic speech recognition.}
    \note[item]<5->{We then move to a different domain, medical coding, and provide a critical review of recent works.}
    \note[item]<6->{Finally, we provide a retrospective study on a machine learning-assisted stroke recognition system for medical helpline calls.}
    \end{frame}

### 2024-03-03 presentation/parts/presentation_hierarchical.tex
WIP

\frame{
    \frametitle{Defining OOD detection}
    \begin{columns}
        \begin{column}{0.5\textwidth}
            Enable models to distinguish the training data distribution $p(\xb)$ from any other distribution $\tilde{p}(\xb)$.
            \vspace{3mm}
            Do this for any given single observation, i.e. answer the question:
            \vspace{3mm}
            \begin{center}
                "Was $\xb$ sampled from $p(\xb)$ or not?"
            \end{center}
        \end{column}
        \begin{column}{0.5\textwidth}
            \begin{figure}[\textwidth]
                \centering
                \begin{tikzpicture}
                        \begin{axis}[
                            name=axistop,
                            domain=0:13,samples=100,
                            height=4.5cm, width=8cm,
                            xtick=\empty,
                            every axis plot/.append style={very thick,smooth}
                        ]
                            \addplot [name path=p1, color=red!50!black] {gauss(5,1)};
                            \addplot [name path=p2, color=black] {gauss(8,1)};
                            \path[name path=axis] (axis cs:0,0) -- (axis cs:10,0);
                            \addlegendentry{$p(\xb)$}
                            \addlegendentry{$\tilde{p}(\xb)$}
                        \end{axis}
                        \path (axistop.south) ++ (0,-0.4cm) coordinate (test);
                        \begin{axis}[
                            name=axisbottom,
                            anchor=north,
                            shift=(test),
                            domain=0:13,samples=100,
                            height=4.5cm, width=8cm,
                            every axis plot/.append style={very thick,smooth}
                        ]
                            \addplot [name path=p1, color=red!50!black] {gauss(4,0.5)};
                            \addplot [name path=p2, color=black] {gauss(9,0.5)};
                            \path[name path=axis] (axis cs:0,0) -- (axis cs:10,0);
                            \addlegendentry{$p(\xb)$}
                            \addlegendentry{$\tilde{p}(\xb)$}
                        \end{axis}
                \end{tikzpicture}
            \end{figure}
        \end{column}
    \end{columns}
}

@@

\frame{
    \frametitle{Out of distribution?}
    \begin{figure}[\textwidth]
        \centering
        \includegraphics[scale=0.6]{figures/dataexamples.pdf}
    \end{figure}
}

\frame{
    \frametitle{Problem and Contributions}
    \begin{itemize}
        \item Deep generative models often fail at OOD detection task when using their likelihood estimate as the score function \cite{nalisnick_deep_2019} by, perhaps surprisingly, assigning \highlight{higher likelihoods to OOD data}.
        \item Contributions:
        \begin{itemize}
            \item We provide evidence that out-of-distribution detection fails due to learned low-level features that generalize across datasets.
            \item We present a new score for OOD detection with hierarchical VAEs that alleviates this issue.
        \end{itemize}  
    \end{itemize}
}

@@

\begin{frame}
    \frametitle{Out-of-distribution detection with hierarchical VAEs}
    \begin{columns}[t]
        \begin{column}{0.4\textwidth}
            \begin{itemize}
                \item Generative models learn to approximate the \textbf{data distribution} $p(\xb)$.
                \item The likelihood of the model given a sample $\xb$ is a measure of how well the model \textbf{explains the data}.
                \item \textbf{Model likelihood} has long been thought of as useful for OOD detection \cite{bishop_novelty_1994}.
            \end{itemize}
        \end{column}
        \begin{column}{0.6\textwidth}
            \begin{overprint}
                \onslide<2>
                \begin{figure}
                    \centering
                    \includegraphics[width=\textwidth]{../graphics/paper_hierarchical/densities-FashionMNIST-test-MNIST-test-bpp-k-0_sohau.pdf}
                \end{figure}
            \end{overprint}
        \end{column}
    \end{columns}
\end{frame}

@@

\frame{
    \frametitle{What is wrong with the ELBO for OOD detection?}
    We can split the ELBO into two terms
    \begin{equation}
        \mathcal{L}(\xb;\theta,\phi) 
        = \mathbb{E}_{q_\phi(\zb|\xb)} \left[ \log \frac{p_\theta(\xb,\zb)}{q_\phi(\zb|\xb)} \right] 
        = \underbrace{\mathbb{E}_{q_\phi(\zb|\xb)}[\log p_\theta(\xb|\zb)]}_{\text{reconstruction likelihood}} - \underbrace{D_{\mathrm{KL}}( q_\phi(\zb|\xb) || p(\zb))}_{\text{regularization penalty}} \ .
    \end{equation}
    The first term is high if the data is well-explained by $\zb$.

    \vspace{3mm}
    The second term we can rewrite as,
    \begin{equation}
        D_{\mathrm{KL}}( q_\phi(\zb|\xb) || p(\zb)) = \mathbb{E}_{q_\phi(\zb|\xb)} \Big[ \textstyle\sum_{i=1}^{L-1} \log \frac{p_\theta(\zb_i|\zb_{i+1})}{q_\phi(\zb_i|\zb_{i-1})} + \log \frac{p_\theta(\zb_L)}{q_\phi(\zb_L|\zb_{L-1})} \Big] \ .
    \end{equation}
    The absolute log-ratios grow with $\mathrm{dim}(\zb_i)$ since the log probability terms are computed by summing over the dimensionality of $\zb_i$.

    \vspace{3mm}
}

@@

    \begin{figure}[\textwidth]
        \centering
        \includegraphics[scale=0.75]{../graphics/paper_hierarchical/feature-correlation-heatmap2.pdf}
        % \caption{Caption}
        % \label{fig:my_label}
    \end{figure}
}

\section{The $\mathcal{L}^{>k}$ likelihood bound}

@@

    This bound uses the conditional prior for the lowest latent variables in the hierarchy.
}

\section{Likelihood ratio}

\frame{
    \frametitle{Likelihood ratios}
    We can use our new bound to compute the score used in a standard likelihood ratio test \cite{buse_likelihood_1982}.
    \begin{equation}\label{eq:llr-as-difference-in-likelihoods}
        LLR^{>k}(\xb) \equiv \mathcal{L}(\xb) - \mathcal{L}^{>k}(\xb) \ .
    \end{equation}
    We can inspect what this likelihood-ratio measures by considering the exact form of our bounds.
    \begin{align}
        \mathcal{L}      &= \log p_\theta(\xb) - D_{\mathrm{KL}}\left( q_\phi(\zb|\xb) || p_\theta(\zb|\xb)\right), \label{eq:likelihoods-as-exact} \\ 
        \mathcal{L}^{>k} &= \log p_\theta(\xb) - D_{\mathrm{KL}}\left( p_\theta(\zb_{\leq }|\zb_{>k}) q_\phi(\zb_{>k}|\xb) || p_\theta(\zb|\xb)\right) \notag \ .
    \end{align}
    In the likelihood ratio the reconstruction terms cancel out and only the KL-divergences from the approximate to the true posterior remain.
    \begin{align}\label{eq:llr-as-kls}
        LLR^{>k}(\xb) &= - D_{\mathrm{KL}}\left( q_\phi(\zb|\xb) || p_\theta(\zb|\xb)\right) \\
                     &\quad + D_{\mathrm{KL}}\left( p_\theta(\zb_{\leq }|\zb_{>k}) q_\phi(\zb_{>k}|\xb) || p_\theta(\zb|\xb)\right) \ . \notag
    \end{align}
}

@@

    Consequently, by importance sampling the ELBO, the associated KL-divergence vanishes and our likelihood ratio reduces to the KL-divergence of $\mathcal{L}^{>k}$.

    \begin{equation}\label{eq:llr-as-kls-iwae-reduced}
        LLR^{>k}_{S}(\xb) \rightarrow D_{\mathrm{KL}}( p(\zb_{\leq }|\zb_{>k}) q(\zb_{>k}|\xb) || p(\zb|\xb)) \ .
    \end{equation}

    $LLR^{>k}_{S}(\xb)$ performs KL-divergence-based OOD detection using top-most latent variables.
}

@@

\frame{
    \frametitle{Results on diverse datasets}
    \begin{columns}
        \begin{column}{0.45\textwidth}
            \hfill
            \begin{table}[t]
                \centering
                \resizebox{0.9\textwidth}{!}{%
                \begin{tabular}{llrrr}
                    \toprule
                    OOD dataset & Metric & AUROC$\uparrow$ & AUPRC$\uparrow$ & FPR80$\downarrow$ \\
                    \midrule
                    \multicolumn{5}{c}{\textbf{Trained on CIFAR10}} \\
                    \midrule
                    SVHN          &  $LLR^{>2}$  &  0.811  &  0.837  &  0.394  \\
                    CIFAR10       &  $LLR^{>1}$  &  0.469  &  0.479  &  0.835  \\
                    \midrule
                    \multicolumn{5}{c}{\textbf{Trained on SVHN}} \\
                    \midrule
                    CIFAR10            &  $LLR^{>1}$       &  $0.939$  &  $0.950$  &  $0.052$  \\
                    SVHN               &  $LLR^{>1}$       &  $0.489$  &  $0.484$  &  $0.799$  \\
                    \bottomrule
                \end{tabular}
                }
            \end{table}
            \hfill
        \end{column}
        \begin{column}{0.55\textwidth}
            \hfill
            \begin{table}[t]
                \centering
                \resizebox{0.9\textwidth}{!}{%
                \begin{tabular}{llrrr}
                    \toprule
                    OOD dataset & Metric & AUROC$\uparrow$ & AUPRC$\uparrow$ & FPR80$\downarrow$ \\
                    \midrule
                    \multicolumn{5}{c}{\textbf{Trained on FashionMNIST}} \\
                    \midrule
                    MNIST                    & $LLR^{>1}$             &  0.986  &  0.987  &  0.011 \\
                    notMNIST                 &  $LLR^{>1}$            &  0.998  &  0.998  &  0.000 \\
                    KMNIST                   &  $LLR^{>1}$            &  0.974  &  0.977  &  0.017 \\
                    Omniglot28x28            &  $LLR^{>2}$            &  1.000  &  1.000  &  0.000 \\
                    Omniglot28x28Inverted    &  $LLR^{>1}$            &  0.954  &  0.954  &  0.050 \\
                    SmallNORB28x28           &  $LLR^{>2}$            &  0.999  &  0.999  &  0.002 \\
                    SmallNORB28x28Inverted   &  $LLR^{>2}$            &  0.941  &  0.946  &  0.069 \\
   {\color{black!60}FashionMNIST}             &  $LLR^{>1}$            &  0.488  &  0.496  &  0.811 \\
                    \midrule
                    \multicolumn{5}{c}{\textbf{Trained on MNIST}} \\
                    \midrule
                    FashionMNIST                   &  $LLR^{>1}$  &  $0.999$  &  $0.999$  &  $0.000$ \\
                    notMNIST                       &  $LLR^{>1}$  &  $1.000$  &  $0.999$  &  $0.000$ \\
                    KMNIST                         &  $LLR^{>1}$  &  $0.999$  &  $0.999$  &  $0.000$ \\
                    Omniglot28x28                  &  $LLR^{>1}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
                    Omniglot28x28Inverted          &  $LLR^{>1}$  &  $0.944$  &  $0.953$  &  $0.057$ \\
                    SmallNORB28x28                 &  $LLR^{>1}$  &  $1.000$  &  $1.000$  &  $0.000$ \\
                    SmallNORB28x28Inverted         &  $LLR^{>1}$  &  $0.985$  &  $0.987$  &  $0.000$ \\
   {\color{black!60}MNIST}                          &  $LLR^{>2}$  &  $0.515$  &  $0.507$  &  $0.792$ \\
                    \bottomrule
                \end{tabular}
                }
            \end{table}
            \hfill
        \end{column}
    \end{columns}
}

\begin{frame}
    \frametitle{Conclusions}
    \begin{itemize}
        \item Key observations: 
        \begin{itemize}
            \item The likelihood of a generative model is not a good score for OOD detection \cite{nalisnick_deep_2019}.
            \item Strong correlations between some latent variables for different datasets.
            \item Reconstructions of OOD data are good when using full approximate posterior.
        \end{itemize}
        \item[\highlight{$\rightarrow$}] Proposed a new score, $LLR^{>k}$, that uses the conditional prior for the top-most latent variables in the hierarchy.
    \end{itemize}
\end{frame}

>>>>

\frame{
    \frametitle{Defining OOD detection}
    \begin{columns}
        \begin{column}{0.5\textwidth}
            Enable models to distinguish the training data distribution $p(\xb)$ from any other distribution $\tilde{p}(\xb)$.
            \vspace{3mm}

            Do this for any given single observation, i.e. answer the question:

            \vspace{3mm}
            \begin{center}
                "Was $\xb$ sampled from $p(\xb)$ or not?"
            \end{center}
        \end{column}
        \begin{column}{0.5\textwidth}
            \begin{figure}[\textwidth]
                \centering
                \begin{tikzpicture}
                        \begin{axis}[
                            name=axistop,
                            domain=0:13,samples=100,
                            height=4.5cm, width=8cm,
                            xtick=\empty,
                            every axis plot/.append style={very thick,smooth}
                        ]
                            \addplot [name path=p1, color=red!50!black] {gauss(5,1)};
                            \addplot [name path=p2, color=black] {gauss(8,1)};
                            \path[name path=axis] (axis cs:0,0) -- (axis cs:10,0);
                            \addlegendentry{$p(\xb)$}
                            \addlegendentry{$\tilde{p}(\xb)$}
                        \end{axis}
                        \path (axistop.south) ++ (0,-0.4cm) coordinate (test);
                        \begin{axis}[
                            name=axisbottom,
                            anchor=north,
                            shift=(test),
                            domain=0:13,samples=100,
                            height=4.5cm, width=8cm,
                            every axis plot/.append style={very thick,smooth}
                        ]
                            \addplot [name path=p1, color=red!50!black] {gauss(4,0.5)};
                            \addplot [name path=p2, color=black] {gauss(9,0.5)};
                            \path[name path=axis] (axis cs:0,0) -- (axis cs:10,0);
                            \addlegendentry{$p(\xb)$}
                            \addlegendentry{$\tilde{p}(\xb)$}
                        \end{axis}
                \end{tikzpicture}
            \end{figure}
        \end{column}
    \end{columns}

    \note[item]{Define OOD}
    \note[item]{In 1D it is easy to see when this is possible and when it is not.}
    \note[item]{In higher dimensions it is not as easy to see.}
}

@@

\frame{
    \frametitle{Out of distribution?}
    \begin{figure}[\textwidth]
        \centering
        \includegraphics[scale=0.6]{figures/dataexamples.pdf}
    \end{figure}

    \note[item]{Datasets can overlap quite a bit in their raw data space.}
}

\begin{frame}
    \frametitle{Out-of-distribution detection with generative models}
    \begin{columns}[t]
        \begin{column}{0.4\textwidth}
            \begin{itemize}
                \item Generative models learn to approximate the \textbf{data distribution} $p(\xb)$.
                \item The likelihood of the model given a sample $\xb$ is a measure of how well the model \textbf{explains the data}.
                \item \textbf{Model likelihood} has long been thought of as useful for OOD detection \cite{bishop_novelty_1994}.
            \end{itemize}
        \end{column}
        \begin{column}{0.6\textwidth}
            \begin{overprint}
                \onslide<2>
                {\footnotesize HVAE trained on FashionMNIST evaluated on FashionMNIST and MNIST.}
                \begin{figure}
                    \centering
                    \includegraphics[width=0.8\textwidth]{../graphics/paper_hierarchical/densities-FashionMNIST-test-MNIST-test-bpp-k-0_sohau.pdf}
                \end{figure}
            \end{overprint}
        \end{column}
    \end{columns}
\end{frame}

% \frame{
%     \frametitle{Problem and Contributions}
%     \begin{itemize}
%         \item Deep generative models often fail at OOD detection task when using their likelihood estimate as the score function \cite{nalisnick_deep_2019} by, perhaps surprisingly, assigning \highlight{higher likelihoods to OOD data}.
%         \item Contributions:
%         \begin{itemize}
%             \item We provide evidence that out-of-distribution detection fails due to learned low-level features that generalize across datasets.
%             \item We present a new score for OOD detection with hierarchical VAEs that alleviates this issue.
%         \end{itemize}  
%     \end{itemize}
% }

@@

@@

\frame{
    \frametitle{What is wrong with the ELBO for OOD detection?}
    We can split the ELBO into two terms
    \begin{equation}
        \mathcal{L}(\xb;\theta,\phi) 
        = \mathbb{E}_{q_\phi(\zb|\xb)} \left[ \log \frac{p_\theta(\xb,\zb)}{q_\phi(\zb|\xb)} \right] 
        = \underbrace{\mathbb{E}_{q_\phi(\zb|\xb)}[\log p_\theta(\xb|\zb)]}_{\text{reconstruction likelihood}} - \underbrace{D_{\mathrm{KL}}( q_\phi(\zb|\xb) \parallel p(\zb))}_{\text{regularization penalty}} \ .
    \end{equation}
    The first term is high if the data is well-explained by $\zb$.

    \vspace{3mm}
    The second term we can rewrite as,
    \begin{equation}
        D_{\mathrm{KL}}( q_\phi(\zb|\xb) \parallel p(\zb)) = \mathbb{E}_{q_\phi(\zb|\xb)} \Big[ \textstyle\sum_{i=1}^{L-1} \log \frac{p_\theta(\zb_i|\zb_{i+1})}{q_\phi(\zb_i|\zb_{i-1})} + \log \frac{p_\theta(\zb_L)}{q_\phi(\zb_L|\zb_{L-1})} \Big] \ .
    \end{equation}
    Since the individual terms are computed by summing over the dimensionality of $\zb_i$, the absolute log-ratios grow with $\mathrm{dim}(\zb_i)$.

    \note[item]{There is a tendency for the lower-most latent variables to be weighted higher than the higher ones.}
}

@@

    \begin{figure}[\textwidth]
        \centering
        \includegraphics[scale=0.75]{../graphics/paper_hierarchical/feature-correlation-heatmap2.pdf}
        % \caption{Caption}
        % \label{fig:my_label}
    \end{figure}

    \note[item]{Strong evidence that the lowest latent variables are generalizing across datasets.}
}

% \section{The $\mathcal{L}^{>k}$ likelihood bound}

@@

    This bound uses the conditional prior for the lowest latent variables in the hierarchy.

    \note[item]{So can we come up with a new bound that does not use the lowest latent variables in the same way?}
    \note[item]{So we could use this for OOD detection (as done in BIVA).}
}

% \section{Likelihood ratio}

\frame{
    \frametitle{Likelihood ratios}
    We can use our new bound to compute the score used in a standard likelihood ratio test \cite{buse_likelihood_1982}.
    \begin{equation}\label{eq:llr-as-difference-in-likelihoods}
        LLR^{>k}(\xb) \equiv \mathcal{L}(\xb) - \mathcal{L}^{>k}(\xb) \ .
    \end{equation}
    We can inspect what this likelihood-ratio measures by considering the exact form of our bounds.
    \begin{align}
        \mathcal{L}      &= \log p_\theta(\xb) - D_{\mathrm{KL}}\left( q_\phi(\zb|\xb) \parallel p_\theta(\zb|\xb)\right), \label{eq:likelihoods-as-exact} \\ 
        \mathcal{L}^{>k} &= \log p_\theta(\xb) - D_{\mathrm{KL}}\left( p_\theta(\zb_{\leq }|\zb_{>k}) q_\phi(\zb_{>k}|\xb) \parallel p_\theta(\zb|\xb)\right) \notag \ .
    \end{align}
    In the likelihood ratio the reconstruction terms cancel out and only the KL-divergences from the approximate to the true posterior remain.
    \begin{align}\label{eq:llr-as-kls}
        LLR^{>k}(\xb) &= - D_{\mathrm{KL}}\left( q_\phi(\zb|\xb) \parallel p_\theta(\zb|\xb)\right) \\
                     &\quad + D_{\mathrm{KL}}\left( p_\theta(\zb_{\leq }|\zb_{>k}) q_\phi(\zb_{>k}|\xb) \parallel p_\theta(\zb|\xb)\right) \ . \notag
    \end{align}

    \note[item]{Write likelihood-ratio using the exact form of the bounds including intractable KL-divergence.}
}

@@

    Consequently, by importance sampling the ELBO, $D_{\mathrm{KL}}\left( q_\phi(\zb|\xb) \parallel p_\theta(\zb|\xb)\right) \rightarrow 0$ and our likelihood ratio reduces to the KL-divergence of $\mathcal{L}^{>k}$.

    \begin{equation}\label{eq:llr-as-kls-iwae-reduced}
        LLR^{>k}_{S}(\xb) \rightarrow D_{\mathrm{KL}}( p(\zb_{\leq }|\zb_{>k}) q(\zb_{>k}|\xb) \parallel p(\zb|\xb)) \ .
    \end{equation}

    $LLR^{>k}_{S}(\xb)$ performs KL-divergence-based OOD detection using top-most latent variables.

    \note[item]{We can use the IWAE bound force the KL-divergence of the original ELBO towards zero.}
    \note[item]{Then, the likelihood ratio reduces to the KL-divergence of $\mathcal{L}^{>k}$.}
    \note[item]{We can hope that using the conditional prior for the lowest latent variables will make the KL-divergence more sensitive to OOD data in the top-most latent variables.}
}

@@

\begin{frame}
    \frametitle{Conclusions}
    \begin{itemize}
        \item Key observations: 
        \begin{itemize}
            \item The likelihood of a generative model is not a good score for OOD detection \cite{nalisnick_deep_2019}.
            \item Strong correlations between some latent variables for different datasets.
            \item Reconstructions of OOD data are good when using full approximate posterior.
        \end{itemize}
        \item[\highlight{$\rightarrow$}] Proposed a likelihood-ratio score, $LLR^{>k}$, that uses the conditional prior for the top-most latent variables in the hierarchy.
        \item[\highlight{$\rightarrow$}] Showed that $LLR^{>k}$ outperforms the standard likelihood score across many different datasets and is competetive with state-of-the-art methods.
    \end{itemize}
\end{frame}
